Black-swan events in animal populations

### 2017-01-07 text/ms/ms_pnas.tex
Complete subfolder

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}ajor surprises (black-swan events) happen more often than expected in
financial, social, and natural systems
\cite{sornette2002, albeverio2006, taleb2007, may2008, sornette2009}.
Massive unpredictable market swings are responsible for the majority of
financial gains and losses \cite{taleb2007}, fatalities from the largest wars
dwarf those from all others \cite{newman2005}, and the frequency of the most
damaging earthquakes has exceeded past expectations \cite{sornette2009}. In
ecological systems,
billions of animals can die at once in mass mortality
events \cite{fey2015},
background rates of global extinction are punctuated by
mass extinction \cite{harnik2012},
and evolution is characterized by
long periods of stasis punctuated by
bursts of speciation \cite{gould1977}.

@@

We found that black-swan dynamics are infrequent but highly
influential: in our dataset they were detected most
frequently for birds (\birdPH\%), mammals (\mammalsPH\%), and
insects (\insectsPH\%) but almost never in fishes (Fig.~2). Black-swan
events were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders.
Accounting for time series length and partially pooling inference across
taxonomic class and order with a hierarchical model, we found stronger evidence
for black swans in insect populations than these statistics suggest---two of
five orders with the highest median probability of heavy tails were insect
orders (Fig.~3A).

\begin{figure}[htb]
\centering
\includegraphics[width=0.28\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (A to D)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel A are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

The majority of our heavy-tailed estimates were robust to alternative
population models, observation error, and choice of Bayesian priors (Supporting Information).
Our conclusions were not systematically altered when we included autocorrelation
in the residuals, modelled population growth rates with or without
density dependence, or modelled the population dynamics as Ricker-logistic
instead of Gompertz (Fig.~S\figalt). Similarly, introducing moderate
observation error (CV = 0.2) only slightly decreased the estimated prevalence
of black-swan events (Fig.~S\figalt), and the strength of the prior on
\(\nu\) had little influence on estimates of black-swan dynamics
(Fig.~S\figaltpriors). Finally, our simulation testing shows that, if
anything, our models underpredict the true magnitude and probability of
heavy-tailed events---especially given that the time series are relatively short,
with a median of only 26
years in our analysis (Figs~S\figsimnu,~S\figsimprob).

% Our results show that weaker and stronger priors have little influence on our
% conclusions about heavy-tailed dynamics (Fig.~\figaltpriors). When the
% data are informative about tail behaviour (i.e. when there is strong evidence
% of low \(\nu\) values, upper-right of Fig.~\figaltpriors), the prior has
% little impact on the estimate of \(\nu\). When the data are less informative
% about \(\nu\) (i.e.\ when there are no or few tail events and time series are
% short or noisy), the prior can pull the estimate of \(\nu\) towards larger or
% smaller values (Fig.~\figaltpriors). The vast majority of the
% populations with Pr\((\nu < 10)\) in the base prior were not altered
% qualitatively by this range of prior strength.

For model fits to the population data, the probability of detecting black-swan
dynamics was positively related to time-series length and negatively related to
magnitude of process noise but not clearly related to population growth rate,
density dependence, or maximum lifespan (Fig.~3B, Fig.~S\figcorrelates).
Longer time-series length was the strongest covariate of observing black-swan
dynamics: there is a \pIncHeavyNThirtyNSixty\ times greater probability of detecting a black-swan
event in populations with 60 time steps than in one with 30 time steps
(Fig.~S\figcorrelates).

\begin{figure}[htb]
\centering
\includegraphics[width=0.26\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{Standardized probabilities and covariates of heavy-tailed
dynamics. (A) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in A indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time-series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (B)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.
}
\label{fig:3}
\end{figure}

The majority of black-swan events (\percBSDown \%) were downward (die-offs)
rather than upward (unexpectedly rapid abundance increases). Of the black-swan
events with published explanations (Table~S2), the majority involved
a combination of multiple factors. For example, a synchronization of
environmental- and predation-mediated population cycles is thought to have
caused a downward black-swan event for a water vole (\emph{Arvicola
terrestris}) population \cite{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For instance, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings
in grey herons (\emph{Ardea cinerea}) in the United
Kingdom \cite{stafford1971} (Fig.~4c). Our analysis finds that the last event
was a combination of two heavy-tailed events in a row and the population took
three times longer to recover than predicted \cite{stafford1971}. Downward black
swans were sometimes followed by upward black swans. For example, during
a period of population crowding and nest shortages, a population of European
shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United
Kingdom declined suddenly following a red tide event in 1968 \cite{potts1980}.
This freed up quality nest sites for first-time breeders, productivity
increased, and the population experienced a rapid upswing in
abundance \cite{potts1980}.

\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{../analysis/skew-fig.pdf}
\caption{Heavy-tailed process noise tends to be downward skewed and
ignoring this can underestimate risk. (A) Illustrations of Student-$t$
distributions with three levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
(B) Posterior density of the skewness parameters aggregated across
populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
(\(\hat{\nu} > 70\)). Approximate mid-values from B are illustrated in
panel A. (C) Example time series of heavy-tailed populations with
different levels of skewness. Red dots highlight likely heavy-tailed events.
Forecasts (grey regions) show median (solid lines) and lower 99\% credible
intervals (dotted lines) of abundance. Black and red lines indicate forecasts
from Gompertz models with lognormal and skew-$t$ process noise, respectively.
}
\label{fig:4}
\end{figure}

Given the prevalence of downwards events, we refit our heavy-tailed models to
measure the degree of skewness \(\gamma\) of the process noise using a skew-$t$
distribution (Fig.~4a), and used these models to make near-term risk forecasts.
Aggregated across populations with strong evidence of heavy tails (median \(\nu
< 10\)), \probDensSkewedForHeavyPops\% of the \(\gamma\)
probability density was below 1, indicating
strong evidence of downwardly skewed process noise (Fig.~4B,
Fig.~S\figskewnu). In contrast, populations that did not have heavy
tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
(Fig.~4B; \percNormPopsNotSkewed\% of 95\% credible intervals overlapped~1). Projecting these
heavy-tailed populations forward five years revealed that assuming standard
normal process noise underestimated risk (99\% lower credible interval of
abundance) by \crashUnderRange-fold (interquartile range; Fig.~4C,
Fig.~S\figskewprojections).

@@

There are a number of caveats when considering the generality of our results.
First, the Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
represents a taxonomically and geographically biased sample of
populations---the longer time series we focused on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Second, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors (Supporting Information).
Third, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.
Finally, our models assume that
the fitted parameters (e.g. productivity and density dependence)
do not vary over time.
Allowing for time-varying parameters
could be an important development for
understanding the dynamics of longer time series.

@@

\matmethods{

\subsection{Data}

We selected abundance time series from the Global Population Dynamics
Database \cite[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (Supporting Information) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated remaining missing values (sensu \cite{brook2006a}).
Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

@@

For the Gompertz model,
we chose weakly informative priors for
all parameters (Fig.~S\figpriors, Supporting Information).
For the degrees of freedom parameter $\nu$,
we chose an exponential prior
with a rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by Ref.~\cite{fernandez1998}. We truncated the
distribution at \(2\) since
below $2$ there can be issues with computational stability.
This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold.
The data must therefore alter the posterior so that
more than 50\% of the posterior probability has $\nu < 10$ for a
population to be declared heavy tailed.
If the data
are uninformative about heavy tails (e.g.~Fig.~1E, H),
and the population model is otherwise a reasonable representation of the dynamics,
the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag any
populations as heavy tailed.

@@

\subsection{Alternative models and simulation testing}
We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence.
This model is equivalent to a random walk with drift
on log abundance, and therefore does not assume
that the time series is stationary.
Third, observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Fourth, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance (Supporting Information).
% Our models also assume that their parameters
% (e.g. parameters representing productivity and density dependence)

We investigated the sensitivity
of our results to weaker and stronger priors (exponential rate parameter $=
0.005, 0.02$) and alternative priors (Fig.~S\figpriors, Supporting Information).
Furthermore, we used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Information).

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from Ref.~\cite{brook2006a}.
Beta regression is useful when response data are on a continuous scale
between zero and one \cite{ferrari2004}. The model was as follows:
\begin{align*}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + \bm{X_i} \bm{\beta}),
  \: \\
  &\quad \, \text{for } i = 1, \dots, 609\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align*}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to
to make the
relationship approximately linear on the logit-transformed response scale;
this matches the way the predictors are visually presented in
Fig.~S\figcorrelates.
All input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in
magnitude \cite{gelman2008c}. We excluded body length as a covariate because it
was highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.

@@

\acknow{We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto,
  two anonymous reviewers, and
members of the Earth to Ocean Group for helpful discussions and comments. We
are grateful to the contributors and maintainers of the Global Population
Dynamics Database and to Compute Canada's WestGrid high-performance computing
resources. Silhouette images were obtained from \url{phylopic.org} under Creative
Commons licenses; sources are listed in the Supporting Information. Funding was
provided by a Simon Fraser University Graduate Fellowship and David H. Smith
Conservation Research Fellowship (S.C.A.), the Natural Sciences and Engineering
Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program
(N.K.D.).
}

>>>>

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}ajor surprises (black-swan events) happen more often than expected in
financial, social, and natural systems
\cite{sornette2002, albeverio2006, taleb2007, may2008, sornette2009}.
Massive unpredictable market swings are responsible for the majority of
financial gains and losses \cite{taleb2007}, fatalities from the largest wars
dwarf those from all others \cite{newman2005}, and the frequency of the most
damaging earthquakes has exceeded past expectations \cite{sornette2009}. In
ecological systems,
background rates of global extinction are punctuated by
mass extinction \cite{harnik2012},
evolution is characterized by
long periods of stasis interrupted by
bursts of speciation \cite{gould1977},
and billions of animals can die at once in mass mortality
events \cite{fey2015}.

@@

We detected black-swan dynamics most
frequently for birds (\birdPH\%), mammals (\mammalsPH\%), and
insects (\insectsPH\%) but almost never in fishes (Fig.~2). Black-swan
events were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders.
Accounting for time series length and partially pooling inference across
taxonomic class and order with a hierarchical model, we found stronger evidence
for black swans in insect populations than these statistics suggest---two of
five orders with the highest median probability of heavy tails were insect
orders (Fig.~3A).

\begin{figure}[htb]
\centering
\includegraphics[width=0.28\textwidth]{fig2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (A to D)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel A are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

The majority of our heavy-tailed estimates were robust to alternative
population models, observation error, and choice of Bayesian priors (SI Appendix).
Our conclusions were not systematically altered when we included autocorrelation
in the residuals, modelled population growth rates with or without
density dependence, or modelled the population dynamics as Ricker-logistic
instead of Gompertz (SI Appendix, Fig.~S\figalt). Similarly, introducing moderate
observation error (CV = 0.2) only slightly decreased the estimated prevalence
of black-swan events (SI Appendix, Fig.~S\figalt), and the strength of the prior on
\(\nu\) had little influence on estimates of black-swan dynamics
(SI Appendix, Fig.~S\figaltpriors). Finally, our simulation testing shows that, if
anything, our models underpredict the true magnitude and probability of
heavy-tailed events---especially given that the time series are relatively short,
with a median of only 26
years in our analysis (SI Appendix, Figs~S\figsimnu,~S\figsimprob).

% Our results show that weaker and stronger priors have little influence on our
% conclusions about heavy-tailed dynamics (SI Appendix, Fig.~\figaltpriors). When the
% data are informative about tail behaviour (i.e. when there is strong evidence
% of low \(\nu\) values, upper-right of SI Appendix, Fig.~\figaltpriors), the prior has
% little impact on the estimate of \(\nu\). When the data are less informative
% about \(\nu\) (i.e.\ when there are no or few tail events and time series are
% short or noisy), the prior can pull the estimate of \(\nu\) towards larger or
% smaller values (SI Appendix, Fig.~\figaltpriors). The vast majority of the
% populations with Pr\((\nu < 10)\) in the base prior were not altered
% qualitatively by this range of prior strength.

For model fits to the population data, the probability of detecting black-swan
dynamics was positively related to time-series length and negatively related to
magnitude of process noise but not clearly related to population growth rate,
density dependence, or maximum lifespan (Fig.~3B, SI Appendix, Fig.~S\figcorrelates).
Longer time-series length was the strongest covariate of observing black-swan
dynamics: there is a \pIncHeavyNThirtyNSixty\ times greater probability of detecting a black-swan
event in populations with 60 time steps than in one with 30 time steps
(SI Appendix, Fig.~S\figcorrelates).

\begin{figure}[htb]
\centering
\includegraphics[width=0.26\textwidth]{fig3.pdf}
\caption{Standardized probabilities and covariates of heavy-tailed
dynamics. (A) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in A indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time-series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (B)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.
}
\label{fig:3}
\end{figure}

The majority of black-swan events (\percBSDown \%) were downward (die-offs)
rather than upward (unexpectedly rapid abundance increases). Of the black-swan
events with published explanations (Table~S2), the majority involved
a combination of factors. For example, a synchronization of
environmental- and predation-mediated population cycles is thought to have
caused a downward black-swan event for a water vole (\emph{Arvicola
terrestris}) population \cite{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For instance, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings
in grey herons (\emph{Ardea cinerea}) in the United
Kingdom \cite{stafford1971} (Fig.~4c). Our analysis finds that the last event
was a combination of two heavy-tailed events in a row and the population took
three times longer to recover than predicted \cite{stafford1971}. Downward black
swans were sometimes followed by upward black swans. For example, during
a period of population crowding and nest shortages, a population of European
shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United
Kingdom declined suddenly following a red tide event in 1968 \cite{potts1980}.
This freed up quality nest sites for first-time breeders, productivity
increased, and the population experienced a rapid upswing in
abundance \cite{potts1980}.

\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{fig4.pdf}
\caption{Heavy-tailed process noise tends to be downward skewed and
ignoring this can underestimate risk. (A) Illustrations of Student-$t$
distributions with three levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
(B) Posterior density of the skewness parameters aggregated across
populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
(\(\hat{\nu} > 70\)). Approximate mid-values from B are illustrated in
panel A. (C) Example time series of heavy-tailed populations with
different levels of skewness. Red dots highlight likely heavy-tailed events.
Forecasts (grey regions) show median (solid lines) and lower 99\% credible
intervals (dotted lines) of abundance. Black and red lines indicate forecasts
from Gompertz models with lognormal and skew-$t$ process noise, respectively.
}
\label{fig:4}
\end{figure}

Given the prevalence of downwards events, we refit our heavy-tailed models to
measure the degree of skewness \(\gamma\) of the process noise using a skew-$t$
distribution (Fig.~4a), and used these models to make near-term risk forecasts.
Aggregated across populations with strong evidence of heavy tails (median \(\nu
< 10\)), \probDensSkewedForHeavyPops\% of the \(\gamma\)
probability density was below 1, indicating
strong evidence of downwardly skewed process noise (Fig.~4B,
SI Appendix, Fig.~S\figskewnu). In contrast, populations that did not have heavy
tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
(Fig.~4B; \percNormPopsNotSkewed\% of 95\% credible intervals overlapped~1). Projecting these
heavy-tailed populations forward five years revealed that assuming standard
normal process noise underestimated risk (99\% lower credible interval of
abundance) by \crashUnderRange-fold (interquartile range; Fig.~4C,
SI Appendix, Fig.~S\figskewprojections).

@@

There are a number of caveats when considering the generality of our results.
First, the Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
represents a taxonomically and geographically biased sample of
populations---the longer time series we focused on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Second, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors (SI Appendix).
Third, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.
Finally, our models assume that
the fitted parameters (e.g. productivity and density dependence)
do not vary over time.
Allowing for time-varying parameters
could be an important development for
understanding the dynamics of longer time series.

@@

\matmethods{
\subsection{Data}

We selected abundance time series from the Global Population Dynamics
Database \cite[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (SI Appendix) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated remaining missing values (sensu \cite{brook2006a}).
Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

@@

For the Gompertz model,
we chose weakly informative priors for
all parameters (SI Appendix, Fig.~S\figpriors).
For the degrees of freedom parameter $\nu$,
we chose an exponential prior
with a rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by Ref.~\cite{fernandez1998}. We truncated the
distribution at \(2\) since
below $2$ there can be issues with computational stability.
This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold.
The data must therefore alter the posterior so that
more than 50\% of the posterior probability has $\nu < 10$ for a
population to be declared heavy tailed.
If the data
are uninformative about heavy tails (e.g.~Fig.~1E, H),
and the population model is otherwise a reasonable representation of the dynamics,
the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag any
populations as heavy tailed.

@@

\subsection{Alternative models and simulation testing}
We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence.
This model is equivalent to a random walk with drift
on log abundance, and therefore does not assume
that the time series is stationary.
Third, observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Fourth, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance (SI Appendix).
% Our models also assume that their parameters
% (e.g. parameters representing productivity and density dependence)

We investigated the sensitivity
of our results to weaker and stronger priors (exponential rate parameter $=
0.005, 0.02$) and alternative priors (SI Appendix, Fig.~S\figpriors).
Furthermore, we used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (SI Appendix).

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from Ref.~\cite{brook2006a}.
Beta regression is useful when response data are on a continuous scale
between zero and one \cite{ferrari2004}. The model was as follows:
\begin{align*}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + \pmb{X_i} \pmb{\beta}),
  \: \\
  &\quad \, \text{for } i = 1, \dots, 609\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align*}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(\pmb{X_i}\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\pmb{\beta}\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to
make the
relationship approximately linear on the logit-transformed response scale;
this matches the way the predictors are visually presented in
SI Appendix, Fig.~S\figcorrelates.
All input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in
magnitude \cite{gelman2008c}. We excluded body length as a covariate because it
was highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.

@@

\acknow{We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto,
  two anonymous reviewers, and
members of the Earth to Ocean Group for helpful discussions and comments. We
are grateful to the contributors and maintainers of the Global Population
Dynamics Database and to Compute Canada's WestGrid high-performance computing
resources. Silhouette images were obtained from \url{phylopic.org} under Creative
Commons licenses; sources are listed in the SI Appendix. Funding was
provided by a Simon Fraser University Graduate Fellowship and David H. Smith
Conservation Research Fellowship (S.C.A.), the Natural Sciences and Engineering
Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program
(N.K.D.).
}
